---
title: "MOFA2: training a model in R"
author:
- name: "Ricard Argelaguet"
  affiliation: "European Bioinformatics Institute, Cambridge, UK"
  email: "ricard@ebi.ac.uk"
- name: "Britta Velten"
  affiliation: "German Cancer Research Center, Heidelberg, Germany"
  email: "b.velten@dkfz-heidelberg.de"
date: "`r Sys.Date()`"

output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{MOFA2: How to train a model in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Load libraries

```{r message=FALSE}
library(data.table)
library(MOFA2)
```

This vignette contains a detailed tutorial on how to train a MOFA model using R. A concise template script can be found [here](https://github.com/bioFAM/MOFA2/blob/master/template_script.R). Many more examples on application of MOFA to various multi-omics data sets can be found [here](https://github.com/bioFAM/MOFA2/tree/master/examples).

# What is MOFA?

MOFA is an **unsupervised statistical framework for the integration of multi-omic data sets**.  
Intuitively, MOFA can be viewed as a versatile and statistically rigorous generalization of principal component analysis (PCA) to multi-omics data. Given several data matrices with measurements of multiple -omics data types on the same or on overlapping sets of samples, MOFA infers an **interpretable low-dimensional representation in terms of a few latent factors** that (hopefully) captures the relevant signal in the input data.  
Effectively MOFA disentangles the sources of variation in the data, identifying the factors that are shared across multiple data modalities from the factors that drive variability in a single data modality.

## Is MOFA appropriate for my data?

MOFA (and factor analysis models in general) are useful to uncover variation in complex data sets that are expected to contain multiple sources of heterogeneity. This requires a relatively large sample size (at least ~15 samples). In addition, MOFA needs the multi-modal measurements to be derived from the same samples. It is fine if you have samples that are missing some data modality, but there has to be a significant degree of matched measurements. 

# Load data

To create a MOFA object you need to specify four dimensions: samples, features, view(s) and group(s). MOFA objects can be created from a wide range of input formats, including:  

- **a list of matrices**: this is recommended for relatively simple data.
- **a long data.frame**: this is recommended for complex data sets with multiple views and/or groups.
- **MultiAssayExperiment**: to connect with Bioconductor objects.
- **Seurat**: for single-cell genomics users. See [this vignette](https://raw.githack.com/bioFAM/MOFA2/master/MOFA2/vignettes/scRNA_gastrulation.html)

## List of matrices

A list of matrices, where each entry corresponds to one view. Samples are stored in columns and features in rows.

Let's simulate some data to start with
```{r}
data <- make_example_data(
  n_views = 2, 
  n_samples = 200, 
  n_features = 1000, 
  n_factors = 10
)[[1]]

lapply(data,dim)
```

Create the MOFA object:
```{r message=FALSE}
MOFAobject <- create_mofa(data)
```

In case you are using the multi-group functionality, the groups can be specified using the `groups` argument as a vector with the group ID for each sample. Keep in mind that this is a rather advanced option that we discourage for MOFA beginners. For more details on how the multi-group inference works, read the [FAQ section](https://github.com/bioFAM/MOFA2#2-faq-on-the-multi-group-functionality) and [check this vignette](https://raw.githack.com/bioFAM/MOFA2/master/MOFA2/vignettes/scRNA_gastrulation.html). 

```{r message=FALSE}
N = ncol(data[[1]])
groups = c(rep("A",N/2), rep("B",N/2))

MOFAobject <- create_mofa(data, groups=groups)
```

## Long data.frame

A long data.frame with columns `sample`, `feature`, `view`, `group` (optional), `value`.  
In my opinion this is the best format for complex data sets with multiple omics and potentially multiple groups of data. Also, there is no need to add rows that correspond to missing data:

```{r }
filepath <- system.file("extdata", "test_data.RData", package = "MOFA2")
load(filepath)

head(dt)
```

Create the MOFA object
```{r }
MOFAobject <- create_mofa(dt)
print(MOFAobject)
```

## Visualise the structure of the data 
```{r  out.width = "80%"}
plot_data_overview(MOFAobject)
```

# Define options 

## Define data options

- **scale_groups**: if groups have different ranges/variances, it is good practice to scale each group to unit variance. Default is `FALSE`
- **scale_views**: if views have different ranges/variances, it is good practice to scale each view to unit variance. Default is `FALSE`
```{r }
data_opts <- get_default_data_options(MOFAobject)
head(data_opts)
```

## Define model options

- **num_factors**: number of factors
- **likelihoods**: likelihood per view (options are "gaussian", "poisson", "bernoulli"). Default is "gaussian".
- **spikeslab_factors**: use spike-slab sparsity prior in the factors? default is `FALSE`.
- **spikeslab_weights**: use spike-slab sparsity prior in the weights? default is `TRUE`.
- **ard_factors**: use ARD prior in the factors? Default is `TRUE` if using multiple groups.
- **ard_weights**: use ARD prior in the weights? Default is `TRUE` if using multiple views.

Please only change the default model options if you are familiar with the underlying mathematical model
```{r }
model_opts <- get_default_model_options(MOFAobject)
model_opts$num_factors <- 10
head(model_opts)
```


## Define train options

- **maxiter**: number of iterations. Default is 1000.
- **convergence_mode**: "fast" (default), "medium", "slow". For exploration, the fast mode is sufficient. For a final model, consider using medium" or even "slow". It will yield sparser weights.
<!-- - **startELBO**: initial iteration to compute the ELBO (the objective function used to assess convergence). -->
<!-- - **freqELBO**: frequency of computations of the ELBO. -->
- **gpu_mode**: use GPU mode? (needs [cupy](https://cupy.chainer.org/) installed and a functional GPU).
- **stochastic**: use stochastic inference? (default is `FALSE`, only use this functionality with very large data sets and in combination with GPUs).
- **verbose**: verbose mode?
- **seed**: random seed (default is 42)

```{r }
train_opts <- get_default_training_options(MOFAobject)
head(train_opts)
```

# Build and train the MOFA object 

Prepare the MOFA object
```{r message=FALSE}
MOFAobject <- prepare_mofa(
  object = MOFAobject,
  data_options = data_opts,
  model_options = model_opts,
  training_options = train_opts
)
```

Train the MOFA model
```{r}
outfile = file.path(getwd(),"model.hdf5")
MOFAobject.trained <- run_mofa(MOFAobject, outfile)
```

If everything is successful, you should observe an output analogous to the following:
```

######################################
## Training the model with seed 1 ##
######################################

Iteration 1: time=0.03, ELBO=-52650.68, deltaELBO=837116.802 (94.082647669%), Factors=10

(...)

Iteration 9: time=0.04, ELBO=-50114.43, deltaELBO=23.907 (0.002686924%), Factors=10

#######################
## Training finished ##
#######################

Saving model in `/Users/ricard/data/mofa2/hdf5/model.hdf5...`r outfile`.
```

# Downstream analysis

This finishes the tutorial on how to train a MOFA object from R. To continue with the downstream analysis, follow [this tutorial](https://raw.githack.com/bioFAM/MOFA2/master/MOFA2/vignettes/downstream_analysis.html)

<details>
  <summary>**Session Info**</summary>
  
```{r}
sessionInfo()
```

</details>
