---
title: "Vignette illustrating the use of SMOFA on simulated data with spatial covariates"
author: "Britta Velten"
output:
  BiocStyle::html_document:
    toc: true
package: MOFA
vignette: >
  %\VignetteIndexEntry{MOFA: Example of SMOFA framework with simulated data (spatial)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, message=FALSE, warning=FALSE}
library(MOFA2)
library(tidyverse)
library(pheatmap)
```

# Spatial data: Simulate an example data set
To illustrate the SMOFA functionality of MOFA2 we simulate a small example data set with 4 different views and two covariates defining a spatial grid using `make_example_data`. The simulation is based on 4 factors, two of which vary smoothly along the covariate (with different lengthscales) and two are independent of the covariate. To mimic a situation where we only have incomplete data on the samples we mask 50% of the samples on average per view.
```{r}
set.seed(2020)

# generate spatial grid and set number of views and samples
sample_cov <- expand_grid(x = 1:10, y =1:10) %>% t()
N <- ncol(sample_cov)
M <- 4

# generate example data
dd <- make_example_data(sample_cov = sample_cov, n_samples = N, n_factors = 4,
                        n_features = 200, n_views = M, lscales = c(5, 2, 0, 0))
```

# SMOFA framework

## Create a MOFA object with covariates
To create a MOFA object we need to specify the training data and the covariates for pattern detection and inference of smooth factors. Here, `sample_cov` is a matrix with samples in columns and covariates in rows. The sample order must match the order in data columns. Alternatively, a data frame can be provided containing one `sample` columns with samples names matching the sample names in the data.
```{r, message=FALSE, warning=FALSE}
sm <- create_mofa(data = dd$data, covariate = dd$sample_cov)
sm

plot_data_overview(sm)
```

## Prepare a MOFA object
Before training, we can specify various options for the model, the training and the data preprocessing. If no options are specified, the model will use the default options. See also `get_default_data_options`, `get_default_model_options` and `get_default_training_options` to have a look at the defaults and change them where required. Here, we drop factors that explain less than 1% of variance in the data starting from 10 factors. We use a grid of 50 points to optimise the lengthscale of each factor and use the moe accurate (though slower) multivariate Z node. To reproduce the results we set a seed for training.
```{r, message=FALSE, warning=FALSE}
train_opt <- get_default_training_options(sm)
train_opt$drop_factor_threshold <- 0.01
train_opt$seed <- 2020
train_opt$maxiter <- 100
model_opts <- get_default_model_options(sm)
model_opts$num_factors <- 4
model_opts$mv_Znode <- TRUE
sm <- prepare_mofa(sm, training_options = train_opt, model_options = model_opts)
```

## Run SMOFA
Now, the MOFA object is ready for training. Using `run_mofa` we can fit the model, which is saves in the file specified as `outfile`. If none is specified the output is only saved in a temporary location.
```{r, message=FALSE, warning=FALSE}
sm <- run_mofa(sm)
```

## Down-stream analysis
### Variance explained per factor
Using `plot_variance_explained` we can explore which factor is active in which view. `plot_factor_cor` shows us whether the factors are correlated.
```{r}
plot_variance_explained(sm)
r <- plot_factor_cor(sm)
```

### Relate factors to the covariate
The MOFA model has learnt lenghtscales for each factor that tell us how smoohtly the factor varys along the covariate. A lengthscale of 0 means that the factor captures variation independent of the covariate, a high value tells us that this factor varys very smoothly along the covariate. In this example, we find two factors that are non-smooth and one smooth factor. Using `plot_factors_vs_cov` we can plot the factors along the spatial coordinates, where we can distinguish smooth and non smooth variation across scace
```{r}
get_lengthscales(sm)
plot_factors_on_cov_2d(sm)
```

We can compare this to the factors that were simulated:
```{r}
df_cov <- get_covariates(sm, as.data.frame = T)
df_cov <- spread(df_cov, key = "covariate", value = "value")

df <- data.frame(dd$Z, df_cov)
df <- gather(df, key = "factor", value = "value", starts_with("simulated_factor"))
ggplot(df, aes(x=covariate_1, y=covariate_2, fill = value)) + 
    geom_tile() + 
    facet_grid( ~ factor) + 
    theme_bw() +
    theme(
      axis.text = element_text(size = rel(0.9), color = "black"), 
      axis.title = element_text(size = rel(1.2), color = "black"), 
      axis.line = element_line(color = "black", size = 0.5), 
      axis.ticks = element_line(color = "black", size = 0.5)
    )
```

### Exploration of weights
As with MOFA(+) we can now look deeper into the meaning of these factors by exploring the weigths or performing feature set enrichment analysis.
```{r}
plot_weights(sm, factors = 4, view = 1)
plot_top_weights(sm, factors = 3, view = 2)
```

## Alternative training option: univariate variational approximation 
A slightly faster version of SMOFA that is more suitable for larger sample size can be obtained by setting `mv_Znode` to `FALSE`. By this, the model uses a faster, more approximate inference that still uses the information provided by the covariate.
```{r, message=FALSE, warning=FALSE}
sm_univ <- create_mofa(data = dd$data, covariate = dd$sample_cov)
model_opts$mv_Znode <- FALSE
sm_univ <- prepare_mofa(sm_univ, training_options = train_opt, model_options = model_opts)
sm_univ <- run_mofa(sm_univ)
```



### Relate factors to the covariate
Again, the SMOFA model has learnt lenghtscales for each factor that tell us how smoohtly the factor varys along the covariate. Here, the factors are slightly less smooth compared to the training with `mv_Znode` set to TRUE.
```{r}
get_lengthscales(sm)
plot_factors_on_cov_2d(sm_univ)
```

# SessionInfo
```{r}
sessionInfo()
```


