---
title: "Vignette illustrating the use of SMOFA on simulated data with a temporal covariate"
author: "Britta Velten"
output:
  BiocStyle::html_document:
    toc: true
package: MOFA
vignette: >
  %\VignetteIndexEntry{MOFA: Example of SMOFA framework with simulated data (temporal)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, message=FALSE, warning=FALSE}
library(MOFA2)
library(tidyverse)
library(pheatmap)
```


# Data: Simulate an example data set
To illustrate the SMOFA functionality of MOFA2 we simulate a small example data set with 4 different views and a single covariate using `make_example_data`. The simulation is based on 4 factors, two of which vary smoothly along the covariate (with different lengthscales) and two are independent of the covariate. To mimic a situation where we only have incomplete data on the samples we mask 50% of the samples on average per view.

```{r}
set.seed(2020)

# number of views and samples
M <- 4
N <- 200

# generate example data with a single group
dd <- make_example_data(n_samples = N, n_features = 500, n_factors = 4,
                        n_views = M, lscales = c(100, 30, 0,0), sample_cov = 1:N)

# simulate some missingness
mask <- sample(1:N, N*0.5*M, replace = TRUE)
mask <- split(mask, 1:M)
dd$data <- lapply(1:M, function(m) {
  mdd <- dd$data[[m]]
  mdd[, mask[[m]]] <- NA
  mdd
  })
```


# SMOFA framework

## Create a MOFA object with covariates
To create a MOFA object for the SMOFA framework we need to specify the training data and the covariates for pattern detection and inerence of smooth factors. Here, `sample_cov` is a matrix with samples in columns and covariates in rows. The sample order must match the order in data columns. Alternatively, a data frame can be provided containing one `sample` columns with samples names matching the sample names in the data.
```{r, message=FALSE, warning=FALSE}
sm <- create_mofa(data = dd$data, sample_cov = dd$sample_cov)
sm

plot_data_overview(sm)
```

## Prepare a MOFA object
Before training, we can specify various options for the model, the training and the data preprocessing. If no options are specified, the model will use the default options. See also `get_default_data_options`, `get_default_model_options` and `get_default_training_options` to have a look at the defaults and change them where required. Here, we drop factors that explain less than 1% of variance in the data starting from 10 factors. We use a grid of 50 points to optimise the lengthscale of each factor and use the moe accurate (though slower) multivariate Z node. To reproduce the results we set a seed for training.
```{r, message=FALSE, warning=FALSE}
train_opt <- get_default_training_options(sm)
train_opt$drop_factor_threshold <- 0
train_opt$seed <- 1234567
train_opt$maxiter <- 100
model_opts <- get_default_model_options(sm)
model_opts$num_factors <- 4
model_opts$mv_Znode <- TRUE
model_opts$GP_factors <- TRUE
sm <- prepare_mofa(sm, training_options = train_opt, model_options = model_opts)
```
## Run MOFA
Now, the MOFA object is ready for training with the SMOFA options. Using `run_mofa` we can fit the model, which is saved in the file specified as `outfile`. If none is specified the output is only saved in a temporary location.
```{r, message=FALSE, warning=FALSE}
sm <- run_mofa(sm)
sm
```

## Down-stream analysis
### Variance explained per factor
Using `plot_variance_explained` we can explore which factor is active in which view. `plot_factor_cor` shows us whether the factors are correlated.
```{r}
plot_variance_explained(sm)
r <- plot_factor_cor(sm)
```
### Relate factors to the covariate
The MOFA model has learnt lenghtscales for each factor that tell us how smoohtly the factor varys along the covariate. A lengthscale of 0 means that the factor captures variation independent of the covariate, a high value tells us that this factor varys very smoothly along the covariate. In this example, we find two factors that are non-smooth and one smooth factor. Using `plot_factors_vs_cov` we can plot the factors along the covariate to visualize this relationship.
```{r}
get_lengthscales(sm)
plot_factors_vs_cov(sm, scale = TRUE)
```

We can compare this to the factors that were simulated:
```{r}
df <- data.frame(dd$Z, get_covariates(sm, as.data.frame = T)) # TODO debug
df <- spread(df, key = "covariate", value = "value")
df <- gather(df, key = "factor", value = "value", starts_with("simulated_factor"))
ggplot(df, aes(x =covariate_1, y = value)) + geom_point() + facet_grid(~factor)
```

### Exploration of weights
As with MOFA(+) we can now look deeper into the meaning of these factors by exploring the weigths or performing feature set enrichment analysis.
```{r}
plot_weights(sm, factors = 4, view = 1)
plot_top_weights(sm, factors = 3, view = 2)
```

## Alternative training option: SMOFA with univariate variational approximation 
A slightly faster version of SMOFA that is more suitbale for larger sample size can be obtained by setting `mv_Znode` to `FALSE`. By this, the model uses a faster, more approximate inference that still uses the information provided by the covariate. 

```{r, message=TRUE, warning=TRUE}
# create new MOFA object
sm_univ <- create_mofa(data = dd$data, sample_cov = dd$sample_cov)

# replace mv_Znode options
model_opts$mv_Znode <- FALSE

# Prepare and train the MOFA object
sm_univ <- prepare_mofa(sm_univ, training_options = train_opt, model_options = model_opts)
sm_univ <- run_mofa(sm_univ)
```


### Relate factors to the covariate
Again, the MOFA model has learnt lenghtscales for each factor that tell us how smoohtly the factor varys along the covariate. Here, the factors are slightly less smooth compared to the training with `mv_Znode` set to TRUE.
```{r}
get_lengthscales(sm_univ)
plot_factors_vs_cov(sm_univ, scale = TRUE)
```

# SessionInfo
```{r}
sessionInfo()
```

